\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-03-15 honors_thesis/journal_paper/paper.tex
Moved discussion to appendix, and other fixes from honors thesis

We also propose another way of making use of these speculative particles,
keeping the best position found instead of the position that PSO actually would
have taken.  We show that for a number of functions, speculative evaluation
gives dramatic improvements over adding additional particles to the swarm.

@@

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait (e.g. a memory cache miss), guess which way the
branch will go and start executing.  If the processor guessed right, execution
is much farther ahead than if it had just idly waited on the branch.  If it
guessed wrong, execution restarts right where it would have been anyway.  This
is called speculative execution or branch prediction.

@@

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
as a parallel adaptation of the Complete topology called Random that has been
shown to simulate the behavior of Complete with far less
communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
be defined as having $n$ neighbors; we use 2 neighbors for all of our
experiments.

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  If a particle is also its own neighbor, as is most
often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
a particle cannot be the source of its $\nbest$ update while not updating its
$\pbest$.  Thus we have $2(n+1)-1$, or $2n+1$, evaluations.
Table~\ref{tab:evals} lists all possible combinations of updates that can be
seen by a particle with two neighbors other than itself, which is the case in
all of the topologies we use.  If all of these speculative positions are
evaluated in parallel along with the original, non-speculative position of the
particle at time $t-1$, then one of the speculative positions has to be the
correct one, and two iterations have been evaluated at once.

@@

Because $2n+1$ speculative evaluations must be performed for each particle, the
choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require
$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
Complete topology.  In that topology, 2 neighbors are chosen at random from the
swarm for each particle at each iteration.  Because a particle is also always
its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report the
results for performing speculative evaluation using both the Ring topology
(where $n$ again is $3$) and the Random topology on a number of common
benchmark functions.

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  First we discuss the
relatively easy case of a centralized parallel PSO algorithm.  In such an
architecture, the master keeps track of all necessary information with only
trivial message passing needed.  Then we discuss the more complicated case of a
distributed algorithm, where each particle is on its own and needs to send and
receive messages to and from other particles.  Finally we discuss the further
complications of a dynamic topology, where a particle's neighbors change from
one iteration to another.

@@

\subsection{Alternate form of message passing}

In some distributed frameworks, like the MapReduce framework that we used in
our experiments, rounds of message passing can be expensive.  The method just
described uses three rounds of message passing for every two iterations
(corresponding to steps 2, 5 and 10 in \alg{centralized}).  Accordingly, we
describe a method that requires only one round of communication for each pair
of iterations, which happens at step 5 of \alg{centralized}.  Many more
messages are needed, but that is sometimes more desirable than synchronizing
all of the machines three times.

We have already described how to use a set of messages to obtain $\p_{t+1}$.
The process is exactly the same to produce each $\n_{t+1}$, requiring the same
messages, except for the neighbor particles instead of the particle itself.  
\fig{messages5} shows graphically how $\noeval{\sset}_{t+3}$ is produced.

@@

Performing speculative evaluation in PSO with a dynamic topology raises a few
issues of its own.  First, with neighbors changing each iteration, messages
that processors pass to their neighbors need to be sent to the correct
neighbors for each iteration.  In centralized algorithms and in distributed
algorithms with several rounds of message passing (the first method described
above), nothing needs to be changed.  However, in the case where processors
reconstruct information about neighboring particles from messages, those
messages need to be sent to the correct neighbors.  A particle cannot simply
send messages to its neighbors' neighbors' neighbors---it needs to send
messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
For every neighbor outward information is sent, the iteration also needs to be
incremented, as information about neighbors' neighbors is used during iteration
$t+1$, and information about neighbors' neighbors' neighbors is used to
reconstruct information about iteration $t+2$.  Also, this method of message
passing again requires the use of random seeds if the topology is random, so
that each processor computes the same neighbors for a particle as all other
processors.

Also, an issue common to all methods of performing speculative evaluation is
that of getting available information from neighbors before generating
speculative children.  In a static topology, at iteration $t$ a particle
already has all of the information about the positions of its neighbors during
iterations $1$ through $t-1$.  If the neighbor finds a better position at 
iteration $t$, the particle updates its neighborhood best, but if it does not,
it still has its old neighborhood best from its neighbors for all previous 
iterations.

@@

This is easily fixed by updating $\noeval{\p}_{t+2}$ with the currently
available information about $\noeval{\nset}_{t+2}$ before producing
$\noeval{\sset}_{t+3}$.  If a particle $\noeval{\p}_{t+2}$ updates its
neighborhood best with the personal bests of $\noeval{\nset}_{t+2}$ before
calculating the next possible positions for $\noeval{\sset}_{t+3}$, there are
still only $2n+1$ possible next positions, and the problem is avoided.

@@

\subsection{Pick the Best Child}

@@

\begin{table}
  \caption{Five Branches to Consider for Statistics}
  \label{tab:branches}
  \centering
  \begin{tabular}{rc|c}
	&$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	1.&No update&No update\\
	2.&Updated&No update\\
	3.&Updated&The particle itself\\
	4.&No update&Some neighbor\\
	5.&Updated&Some neighbor\\
  \end{tabular}
\end{table}

\begin{table}
  \caption{Branch Statistics in PSO}
  \label{tab:stats}
  \centering
  \begin{tabular}{c|c|c|c|c|c|c}
	Topology&Function&1&2&3&4&5\\
	\hline
	Ring&Sphere&0.530&0.093&0.114&0.202&0.062\\
	&Griewank&0.517&0.084&0.122&0.207&0.070\\
	&Rastrigin&0.495&0.048&0.146&0.213&0.099\\
	&Rosenbrock&0.513&0.074&0.129&0.211&0.073\\
	\hline
	Random&Sphere&0.667&0.119&0.026&0.156&0.031\\
	&Griewank&0.690&0.109&0.025&0.149&0.027\\
	&Rastrigin&0.819&0.055&0.015&0.100&0.010\\
	&Rosenbrock&0.742&0.077&0.022&0.140&0.018\\
	\hline
	Complete&Sphere&0.319&0.092&0.002&0.451&0.135\\
	&Griewank&0.353&0.084&0.002&0.441&0.119\\
	&Rastrigin&0.477&0.067&0.002&0.382&0.070\\
	&Rosenbrock&0.353&0.034&0.003&0.544&0.066\\
  \end{tabular}
\end{table}

@@

Another possibility is to pick the best child, as described above.  This
ignores the fact that the branch might have been wrong; we do not care, because
we just want the child that had the best value, anyway.  In almost all of the
experiments that we ran, it turned out that picking the best child performed
better than keeping the particle back an iteration.

@@

>>>>

We show that while speculative evaluation in PSO is an intriguing idea, its
usefulness is limited unless some slight modifications are made to the PSO
algorithm.  When we relax the requirement that speculative evaluation exactly
reproduce the behavior of the original PSO, we can make improvements such as
using all of the extra speculative particles, instead of throwing away all but
one of them, or speculating several iterations ahead instead of just one.
Using these algorithmic modifications, we see dramatic improvements on a number
of functions over simply adding additional particles to the swarm.

@@

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait (e.g. a memory cache miss), guess which way the
branch will go and start executing.  If the processor guesses right, execution
is much farther ahead than if it had just idly waited on the branch.  If it
guesses wrong, execution restarts right where it would have been anyway.  This
is called speculative execution or branch prediction.

@@

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies, a particle is also a
neighbor to itself, in that its own position is considered when updating the
particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
particle $i$ has three neighbors: $i-1$, $i$, and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.  In this paper we use these
topologies, as well as a parallel adaptation of the Complete topology, called
Random, that has been shown to simulate the behavior of Complete with far less
communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
randomly picks two other particles to share information with at each iteration,
along with itself.  Thus in both the Ring and the Random topologies, all
particles have 3 neighbors.

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  Because a particle is also its own neighbor, one
of the $2(n+1)$ speculative positions can be eliminated, as a particle cannot
be the source of its $\nbest$ update while not updating its $\pbest$.  Thus we
have $2(n+1)-1$, or $2n+1$, evaluations.  Table~\ref{tab:evals} lists all
possible combinations of updates that can be seen by a particle with two
neighbors other than itself, which is the case in all of the topologies we use.
If all of these speculative positions are evaluated in parallel along with the
original, non-speculative position of the particle at time $t-1$, then one of
the speculative positions has to be the correct one, and two iterations have
been evaluated at once.

@@

Because $2n+1$ speculative evaluations must be performed for each particle, the
choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require
$O(p^2)$ speculative evaluations per iteration, where $p$ is the swarm size.
Clearly it is much more desirable to have a sparse topology, where $O(np)$ is
much smaller than $O(p^2)$.  However, some functions are better optimized with
the Complete topology and the quick spread of information it entails than with
sparse topologies.  Accordingly, we use the Random topology described
in~\cite{mcnabb-cec09}, which has been shown to effectively simulate the
Complete topology.  In~\sec{results} we report the results for performing
speculative evaluation using both the Ring topology and the Random topology on
a number of common benchmark functions.

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  First we discuss the
relatively easy case of a centralized parallel PSO algorithm with a master
computer and many slaves.  In such an architecture, the master keeps track of
all necessary information with only trivial message passing needed.  Then we
discuss the more complicated case of a distributed algorithm, where each
particle is on its own and needs to send and receive messages to and from other
particles.  Finally we discuss the further complications of a dynamic topology,
where a particle's neighbors change from one iteration to another.

@@

In some distributed frameworks, like the MapReduce framework that we used in
our experiments, rounds of message passing can be expensive.  The method just
described uses three rounds of message passing for every two iterations
(corresponding to steps 2, 5 and 10 in \alg{centralized}).  It is possible to
perform speculative evaluation in PSO with only one round of communication per
two iterations.  However, the methodology is tedius and is not the focus of
this paper, so we defer its description to the Appendix.

@@

Performing speculative evaluation in PSO with a dynamic topology raises a
sticky issue of its own.  In a static topology, at iteration $t$ a particle
already has all of the information about the positions of its neighbors during
iterations $1$ through $t-1$.  If the neighbor finds a better position at
iteration $t$, the particle updates its neighborhood best, but if it does not,
it still has its old neighborhood best from its neighbors for all previous
iterations.

@@

This is easily fixed by updating each new particle $\noeval{\p}_{t+2}$ with the
currently available information about its neighbors $\noeval{\nset}_{t+2}$
before producing its children $\noeval{\sset}_{t+3}$.  If a particle
$\noeval{\p}_{t+2}$ updates its neighborhood best with the personal bests of
$\noeval{\nset}_{t+2}$ before calculating the next possible positions for
$\noeval{\sset}_{t+3}$, there are still only $2n+1$ possible next positions,
and the problem is avoided.

@@

\subsection{Pick the Best Child}
\label{sec:pickbest}

@@

\begin{table}[h]
  \caption{Five Branches to Consider for Statistics}
  \label{tab:branches}
  \centering
  \begin{tabular}{rc|c}
	&$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	1.&No update&No update\\
	2.&Updated&No update\\
	3.&Updated&The particle itself\\
	4.&No update&Some neighbor\\
	5.&Updated&Some neighbor\\
  \end{tabular}
\end{table}

\begin{table}[h]
  \caption{Branch Statistics in PSO}
  \label{tab:stats}
  \centering
  \begin{tabular}{c|c|c|c|c|c|c}
	Topology&Function&1&2&3&4&5\\
	\hline
	\hline
	Ring&Sphere&53.0\%&9.3\%&11.4\%&20.2\%&6.2\%\\
	&Griewank&51.7\%&8.4\%&12.2\%&20.7\%&7.0\%\\
	&Rastrigin&49.5\%&4.8\%&14.6\%&21.3\%&9.9\%\\
	&Rosenbrock&51.3\%&7.4\%&12.9\%&21.1\%&7.3\%\\
	\cline{2-7}
	&Average&{51.3\%}&{7.5\%}&{12.8\%}&{20.8\%}&
	{7.6\%}\\
	\hline
	\hline
	Random&Sphere&66.7\%&11.9\%&2.6\%&15.6\%&3.1\%\\
	&Griewank&69.0\%&10.9\%&2.5\%&14.9\%&2.7\%\\
	&Rastrigin&81.9\%&5.5\%&1.5\%&10.0\%&1.0\%\\
	&Rosenbrock&74.2\%&7.7\%&2.2\%&14.0\%&1.8\%\\
	\cline{2-7}
	&Average&{73.0\%}&{9.0\%}&{2.2\%}&{13.6\%}&
	{2.2\%}\\
	\hline
	\hline
	Complete&Sphere&31.9\%&9.2\%&0.2\%&45.1\%&13.5\%\\
	&Griewank&35.3\%&8.4\%&0.2\%&44.1\%&11.9\%\\
	&Rastrigin&47.7\%&6.7\%&0.2\%&38.2\%&7.0\%\\
	&Rosenbrock&35.3\%&3.4\%&0.3\%&54.4\%&6.6\%\\
	\cline{2-7}
	&Average&{37.6\%}&{6.9\%}&{.2\%}&{45.5\%}&
	{9.8\%}\\
	\hline
	\multicolumn{7}{l}{The numbers in the column headings refer to branch
	numbers in Table~\ref{tab:branches}.}
  \end{tabular}
\end{table}

@@

Another possibility is to pick the best child, as described in \sec{pickbest}.
This ignores the fact that the branch might have been wrong; we do not care,
because we just want the child that had the best value, anyway.  In almost all
of the experiments that we ran, it turned out that picking the best child
performed better than keeping the particle back an iteration.

@@

\appendix[Alternate form of message passing]

Here we describe a method that requires only one round of communication for
each pair of iterations, which happens at step 5 of \alg{centralized}.  Many
more messages are needed, but that is sometimes more desirable than
synchronizing all of the machines three times.

We have already described how to use a set of messages to obtain $\p_{t+1}$.
The process is exactly the same to produce each $\n_{t+1}$, requiring the same
messages, only for the neighbor particles instead of the particle itself.
\fig{messages5} shows graphically how $\noeval{\sset}_{t+3}$ is produced.

One more issue arises when dealing with dynamic topologies.  With neighbors
changing each iteration, messages that processors pass to their neighbors need
to be sent to the correct neighbors for each iteration.  In centralized
algorithms and in distributed algorithms with several rounds of message passing
(the first method described above), nothing needs to be changed.  However, in
the case where processors reconstruct information about neighboring particles
from messages, those messages need to be sent to the correct neighbors.  A
particle cannot simply send messages to its neighbors' neighbors'
neighbors---it needs to send messages to its iteration $t$ neighbors' iteration
$t+1$ neighbors, and so on.  For every neighbor outward information is sent,
the iteration also needs to be incremented, as information about neighbors'
neighbors is used during iteration $t+1$, and information about neighbors'
neighbors' neighbors is used to reconstruct information about iteration $t+2$.
Also, this method of message passing again requires the use of random seeds if
the topology is random, so that each processor computes the same neighbors for
a particle as all other processors.
